Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-20 modelling/modelling.tex
Proofread

Proposed models for the regeneration process were in the form of non-linear ordinary differential equations (ODE). The only independent variable is post-lesion time $t$. The equations consist of four derivatives of dependent variables (number of cells, mRNA expression of cytokines) with respect to $t$. The equations were mostly an interpretation of the proposed interaction maps, with terms correspond to interaction paths.

@@

Our models were built based on the existing experimental findings from Tsarouchas et al. \cite{ref:Tsarouchas}. The available measurement data included the numbers of three kinds of cells (neutrophil $N$, macrophage $\Phi$ and microglia) and the relative mRNA expression of four kinds of cytokines (il-1$\beta$, tnf-$\alpha$, tgf-$\beta$1a and tgf-$\beta$3). As proposed by Tsarouchas et al., neutrophil and macrophage play essential roles in controlling the il-1$\beta$ and tnf-$\alpha$ mediated inflammation. Our initial models focused on the changes of four dependent variables $N$, $\Phi $, $\beta$ (for il-1$\beta$) and $\alpha$ (for tnf-$\alpha$) that are most important and determinative in the regeneration. $N$ and $\Phi$ is of the unit `number of cells', $\beta$ and $\alpha$ is recorded as `relative mRNA expression' (compared to the value at the initial time point, i.e. 0 hpl) and considered unitless.

It was noted that the variance of the measured data is relatively high. The summary statistic was the mean of measurement data at each time point, assuming that measured data was Gaussian-like distributed. To validate this, the distribution of the measured data points was plotted. The result was that at most time points the measurement values are Gaussian-distributed, although some distributions were skewed. One abnormal distribution was observed at time point 120 hpl for macrophage where there are two concentrations. Despite this, the mean value could summarise most data and thus was still used as the target summary data. The mean value at each time point is shown in Figure \ref{fig:obs_data}, which is our target trajectories for the models to fit.

@@

Five models in total are proposed according to different hypotheses. At first, our tests and implementations of ABC SMC for parameter estimations used only the basic model for developing propose. After the parameter inference framework was built and tested, more models were proposed in order to calibrate and improve the basic model such that the observed regeneration process can be better represented.

All these models assumed that the involved interactions are within two kinds of cells (neutrophil and macrophage) and two kinds of cytokines (il-1$\beta$ and tnf-$\alpha$), and used the data presented in Figure \ref{fig:obs_data} for the inference task. Interactions or influences from other cells or cytokines are not considered as there might not be corresponding data available. If such data or experimental findings are available in the future, we can consider them by then as long as these interactions are indeed necessary for our model.

@@

A preliminary model (Eqn. \ref{eq:model1}) was proposed according to \cite{ref:Tsarouchas} and then mainly used to build and test the code of inference framework. An interaction map of the model is shown in Figure \ref{fig:m1}. This model is a simplification of the process described in \cite{ref:Tsarouchas} (Figure \ref{fig:map}) with some minor interactions ignored. This model included 12 parameters, all of which should be positive real numbers. A description of these parameters can be found in Table \ref{table:m1}.

@@

\end{figure}
\begin{align}
    \label{eq:model1}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_N+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\lambda_\Phi+\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}
It was assumed that there are negative feedbacks the cells and cytokines, i.e. overcrowding inhibits the increasing rates. As sources, macrophage produces tnf-$\alpha$ and neutrophil produces il-1$\beta$; however, macrophage was assumed to have a negative effect on the il-1$\beta$, term $1+i_{\beta\Phi}\Phi$ in Eqn. \ref{eq:model1} was used to represent this. Despite the self-driven increase of immune cells, il-1$\beta$ was assumed to promote the increase of both cells. Neutrophil was also assumed to be inhibited by the total immune cells presented in the lesion site, the term $\nu_{N\Phi}N\Phi$ was used to denote this.

\begin{table}[t]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter            & Definition                                       & Units                    \\ [0.5ex]
        \hline\hline
        $\lambda_N$          & Self-increase rate of neutrophil                 & $cell/h$                 \\
        $\kappa_{N\beta}$    & Promoting effect coefficient by il-1$\beta$      & $cell/h$                 \\
        $\mu_N$              & Coefficient of negative feedback of $N$          & $h^{-1}$                 \\
        $\nu_{N\Phi}$        & Coefficient of inhibition of both $N$ and $\Phi$ & $cell^{-1}\cdotp h^{-1}$ \\
        \hline
        $\lambda_\Phi$       & Self-increase rate of macrophage                 & $cell/h$                 \\
        $\kappa_{\Phi\beta}$ & Promoting effect coefficient by il-1$\beta$      & $cell/h$                 \\
        $\mu_\Phi$           & Coefficient of negative feedback of $\Phi$       & $h^{-1}$                 \\
        \hline
        $s_{\beta N}$        & Production rate from $N$                         & $cell^{-1}\cdotp h^{-1}$ \\
        $i_{\beta\Phi}$      & Coefficient of inhibition to the production      & $cell^{-1}$              \\
        $\mu_\beta$          & Coefficient of negative feedback of $\beta$      & $h^{-1}$                 \\
        \hline
        $s_{\alpha\Phi}$     & Production rate from $\Phi$                      & $cell^{-1}\cdotp h^{-1}$ \\
        $\mu_\alpha$         & Coefficient of negative feedback of $\alpha$     & $h^{-1}$                 \\
        \hline
    \end{tabular}
    \caption{Parameters introduced in the basic model (model 1) [REMOVE unit]}
    \label{table:m1}
\end{table}

@@

As the observed data indicated, this dynamic system has a steady-state where after a long time ($\geq 120$ hpl), the inflammation is resolved, and immune cells should not be present at the injury site any more. Regarding this, the parameter for self-increase rate for the macrophage and neutrophil, i.e. $\lambda_\Phi$ and $\lambda_N$  cannot be constant. An exponentially decaying $\lambda$ term is introduced in the new model (model 2, Eqn. \ref{eq:model2}). Also, the inhibition of il-1$\beta$ production, i.e. the term $i_{\beta\Phi}\Phi$ is considered to be ignored for a more simple model, in which case the relative expression of il-1$\beta$ is only affected by the number of neutrophils and the negative feedback from itself. This case corresponds to model 3, written as Eqn. \ref{eq:model3}.

@@

As stated, our models are largely based on the experimental evidence from Tsarouchas et al.\cite{ref:Tsarouchas}. Some hypothesised interactions may differ from a real case, and there may exist undiscovered mechanisms. These could lead to model misspecification, where the model could fit the data but with wrong interaction paths, or the model can hardly fit the data and recover some local features observed in the data.

>>>>

Proposed models for the regeneration process were in the form of non-linear ordinary differential equations (ODE). The only independent variable is post-lesion time $t$. The equations consist of four derivatives of dependent variables (numbers of two kinds of cells, mRNA expression of two types of cytokines) with respect to $t$. The equations were mostly an interpretation of the proposed interaction maps, with terms correspond to interaction paths.

@@

Our models were built based on the existing experimental findings from Tsarouchas et al. \cite{ref:Tsarouchas}. The available measurement data included the numbers of three kinds of cells (neutrophil $N$, macrophage $\Phi$ and microglia) and the relative mRNA expression of four kinds of cytokines (il-1$\beta$, tnf-$\alpha$, tgf-$\beta$1a and tgf-$\beta$3). As proposed by Tsarouchas et al., neutrophil and macrophage play essential roles in controlling the il-1$\beta$ and tnf-$\alpha$ mediated inflammation. Our initial models focused on the changes of four dependent variables $N$, $\Phi $, $\beta$ (for il-1$\beta$) and $\alpha$ (for tnf-$\alpha$) that are most important and determinative in the regeneration. $N$ and $\Phi$ is of the unit `number of cells', $\beta$ and $\alpha$ is recorded as `relative mRNA expression' (compared to the value at the initial time point 0 hpl) and considered unitless.

It is noted that the variance of the measured data is relatively high. The summary statistic was the mean of measurement data at each time point, assuming that measured data was Gaussian-like distributed. To validate this, the distributions of the measured values was plotted. The result was that at most time points the measurement values were Gaussian-distributed, although some distributions were skewed. One abnormal distribution was observed at time point 120 hpl for macrophage where there were two concentrations. Despite this, the mean value could summarise most data and thus was still used as the target summary data. The mean value at each time point is shown in Figure \ref{fig:obs_data}. It is our target trajectories for the models to fit.

@@

Five models in total were proposed according to different hypotheses. At first, our tests and implementations of ABC SMC for parameter estimations used only the basic model for developing propose. After the parameter inference framework was built and tested, more models were proposed in order to calibrate and improve the basic model such that the observed regeneration process can be better represented.

All these models assumed that the involved interactions are within two kinds of cells (neutrophil and macrophage) and two kinds of cytokines (il-1$\beta$ and tnf-$\alpha$), and used the data presented in Figure \ref{fig:obs_data} for the inference task. Interactions or influences from other cells or cytokines were not considered as there might not be corresponding data available. If such data or experimental findings are available in the future, we can consider them by then as long as these interactions are indeed necessary for our model.

@@

A preliminary model (Eqn. \ref{eq:model1}) was proposed according to \cite{ref:Tsarouchas} and then mainly used to build and test the code for the inference framework. An interaction map of the model is shown in Figure \ref{fig:m1}. This model is a simplification of the process described in \cite{ref:Tsarouchas} (Figure \ref{fig:map}) with some minor interactions ignored. This model includes 12 parameters, all of which should be positive real numbers. A description of these parameters can be found in Table \ref{table:m1}.

@@

\end{figure}
\begin{align}
    \label{eq:model1}
    \begin{split}
        &\frac{\mathrm{d} N}{\mathrm{d} t}=\lambda_N+\kappa_{N\beta}\beta-\mu_NN-\nu_{N\Phi}N\Phi\\
        &\frac{\mathrm{d} \Phi}{\mathrm{d} t}=\lambda_\Phi+\kappa_{\Phi\beta}\beta-\mu_\Phi\Phi\\
        &\frac{\mathrm{d} \beta}{\mathrm{d} t}=\frac{s_{\beta N}N}{1+i_{\beta\Phi}\Phi}-\mu_\beta\beta\\
        &\frac{\mathrm{d} \alpha}{\mathrm{d} t}=s_{\alpha\Phi}\Phi-\mu_\alpha\alpha
    \end{split}
\end{align}
It was assumed that there are negative feedbacks for all the cells and cytokines, i.e. overcrowding inhibits the increasing rate. As sources, macrophage produces tnf-$\alpha$ and neutrophil produces il-1$\beta$; however, macrophage was assumed to have a negative effect on the il-1$\beta$; term $1+i_{\beta\Phi}\Phi$ in Eqn. \ref{eq:model1} was used to represent this. Despite the self-driven increase of immune cells, il-1$\beta$ was assumed to promote the increase of both cells. Neutrophil was also assumed to be inhibited by the numbers of immune cells presented in the lesion site; term $\nu_{N\Phi}N\Phi$ was used to denote this.

\begin{table}[t]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter            & Definition                                       & Units                    \\ [0.5ex]
        \hline\hline
        $\lambda_N$          & Self-increase rate of neutrophil                 & $cell/h$                 \\
        $\kappa_{N\beta}$    & Promoting effect coefficient by il-1$\beta$      & $cell/h$                 \\
        $\mu_N$              & Coefficient of negative feedback of $N$          & $h^{-1}$                 \\
        $\nu_{N\Phi}$        & Coefficient of inhibition of both $N$ and $\Phi$ & $cell^{-1}\cdotp h^{-1}$ \\
        \hline
        $\lambda_\Phi$       & Self-increase rate of macrophage                 & $cell/h$                 \\
        $\kappa_{\Phi\beta}$ & Promoting effect coefficient by il-1$\beta$      & $cell/h$                 \\
        $\mu_\Phi$           & Coefficient of negative feedback of $\Phi$       & $h^{-1}$                 \\
        \hline
        $s_{\beta N}$        & Production rate from $N$                         & $cell^{-1}\cdotp h^{-1}$ \\
        $i_{\beta\Phi}$      & Coefficient of inhibition to the production      & $cell^{-1}$              \\
        $\mu_\beta$          & Coefficient of negative feedback of $\beta$      & $h^{-1}$                 \\
        \hline
        $s_{\alpha\Phi}$     & Production rate from $\Phi$                      & $cell^{-1}\cdotp h^{-1}$ \\
        $\mu_\alpha$         & Coefficient of negative feedback of $\alpha$     & $h^{-1}$                 \\
        \hline
    \end{tabular}
    \caption{Parameters introduced in the basic model (model 1)}
    \label{table:m1}
\end{table}

@@

As the observed data indicates, this dynamic system has a steady-state where after a long time ($\geq 120$ hpl), the inflammation is resolved, and immune cells should not be present at the injury site any more. Regarding this, the parameter for self-increase rate of macrophage and neutrophil, i.e. $\lambda_\Phi$ and $\lambda_N$  cannot be constant. An exponentially decaying $\lambda$ term is introduced in the new model (model 2, Eqn. \ref{eq:model2}). Also, the inhibition of il-1$\beta$ production, i.e. the term $i_{\beta\Phi}\Phi$ is considered to be ignored for a more simple model, in which case the relative expression of il-1$\beta$ is only affected by the number of neutrophils and the negative feedback from itself. This case corresponds to model 3, written as Eqn. \ref{eq:model3}.

@@

As stated, our models are largely based on the experimental evidence from Tsarouchas et al.\cite{ref:Tsarouchas}. Some hypothesised interactions may differ from a real case, and there may exist undiscovered mechanisms. These could lead to model misspecification, where the model could fit the data but with wrong interaction paths, or the model can hardly fit the data or recover some local features observed in the data.

### 2020-08-20 performance/performance.tex
Proofread

For HPC systems like Cirrus, \verb|pyabc| uses \verb|multiprocessing| for multi-core parallel sampling. By default, if the number of cores is not specified, it will automatically read the number of available cores and use them all. Cirrus has a 36-core CPU which support hyperthreading, such that the maximal number of cores available to \verb|multiprocessing| is 72.

The experiment was executed on Cirrus, using 8, 16, 24, 36, 54 and 72 cores respectively. Each run was repeated ten times and the average execution time, required sampling numbers are recorded. Hyperthreading was enabled when using 54 and 72 cores. The access to the node that contains computation cores is exclusive, such that the execution would not be affected by other programs of operations.

The implementation of ABC SMC in \verb|pyabc| enables the parallelisation of sampling, which is the most time-consuming part. The rest part of the program is mostly not parallelised, e.g. database I/O and reductions operations. The sampling process involves sampling, perturbation and test of the acceptance criteria, which are computation-intensive.

@@

A high variance of total required samples was observed in the scaling-up experiments: some single runs required much more samples to finish 20 populations, where they were supposed to have relatively similar average required samples such that we can compare the performance directly via speedup or efficiency.

Due to the variant total required samples drawn our attention to the per-second performance. Although the execution time was not improved by using 54 or 72 cores, the required number of samples were higher than that of $p=36$, which indicates slight improvements. However, the high variance made us hard to gave statistically significant conclusions on the per-second performance.

@@

The variant total required samples for different runs under the same setting was most likely the result of different threshold evolve path. Median epsilon schedule and randomness in the sampling process resulted in different sequences of thresholds for different runs, which means different repeated runs may have different convergency path \cite{threshold}. For example, Some runs temporarily converge to a local optimum, while the target epsilon in the next generation cannot possibly be reached if the new particles are still concentrated in the local optimum. In this case, a huge amount of sample will be drawn to find enough accepted particles, and the new generation will be able to jump out of the local optimum and have different concentrations.

@@

To illustrate this, Two independent run, namely Run A and Run B, were chosen from the experiment of $p=24$ and visualised in Figure \ref{fig:local_modes} and Figure \ref{fig:local_para}. These two runs used identical options of the algorithm. Figure \ref{fig:local_modes} (a) shows the comparison of the required samples, where Run A converged to a local mode, and in population 11 took much more sample tires to move away from the local optimum (Figure \ref{fig:local_modes} (b)). Figure \ref{fig:local_para} compares the posterior density distribution before (t=9, cyan) and after (t=13, red) the jumping away: significant movement of concentrations in posterior density distributions are observed, especially in parameters from ODE 1, 2 and 4. An interning movement is that there are two concentrations in $\kappa_{\Phi\beta}$ and $\mu_\Phi$ when t=9, and the majority of particles are concentrated in the second concentration with high parameter values; when t=13, the second concentration distribution disappears and all particles are concentrated near zero.

@@

By looking into the code, some reasons can be found. There exist some `cheap' particles. If at the first time points, a variable reached unacceptable high value (regarded as \verb|Inf| in \verb|Python|), the ODE solver cannot perform further integration but result in a data full of \verb|NaN|s. Such particles cost less time than an usual particle (where values at 120 data points are explicitly calculated and then compared with observed data). Local mode, e.g. generation 11, might take large amount of these `cheap' particles. Also, generation 1 draws samples directly from prior without fitting and using kernels, so it is much faster. The huge amount of time observed in generation 2 is believed to be a result of some initialisation steps in the ABC SMC algorithm.

To conclude, the performance experiments revealed reasonable scaling-up behaviour when the \verb|Python| processes did not exceed number of physical cores. Local modes lead to high variance in total required number of samples, and the relationship between required sample numbers and required execution time are not ideally linear; they all made further detailed analysis difficult.

Additionally, by observing CPU frequency and load curve in-time while running the program, we found that the parallelised part is limited to the sampling, calculation and criterion test process. Some reduction calculations and preparation work for the next generation (e.g. calculation of the new epsilon values, normalisation of data, fit of multivariate kernel) are serial. The database I/O is also serial.  It is noticed that the program took no advantage of graphics card, as \verb|multiprocessing| along can only exploit the resources inside CPU.

>>>>

For HPC systems like Cirrus, \verb|pyabc| uses \verb|multiprocessing| for multi-core parallel sampling. By default, if the number of cores is not specified, it will automatically read the number of available cores and use them all. Cirrus has a 36-core CPU which supports hyperthreading, such that the maximal number of cores available to \verb|multiprocessing| is 72.

The experiment was executed on Cirrus, using 8, 16, 24, 36, 54 and 72 cores respectively. Each run was repeated ten times. The average execution time and required number of samples are recorded. Hyperthreading was enabled when using 54 and 72 cores. The access to the node that contains computation cores is exclusive, such that the execution would not be affected by other programs of operations.

The implementation of ABC SMC in \verb|pyabc| enables the parallelisation of sampling process, which is the most time-consuming part. The rest part of the program is mostly not parallelised, e.g. database I/O and reductions operations. The sampling process involves sampling, perturbation and test of the acceptance criteria, which are computation-intensive.

@@

A high variance of total required number of samples was observed in the scaling-up experiments: some single runs required much more samples to finish 20 populations, where they were supposed to have relatively similar average required samples such that we can compare the performance directly via speedup or efficiency.

The variant total required number of samples drawn our attention to the per-second performance. Although the execution time was not improved by using 54 or 72 cores, the required number of samples were higher than that of $p=36$, which indicates slight improvements. However, the high variance made us hard to gave statistically significant conclusions on the per-second performance.

@@

The variant total required number of samples for different runs under the same setting was most likely the result of different threshold evolve path. Median epsilon schedule and randomness in the sampling process resulted in different sequences of thresholds for different runs, which means different repeated runs may have different convergency path \cite{threshold}. For example, Some runs temporarily converge to a local optimum, while the target epsilon in the next generation cannot possibly be reached if the new particles are still concentrated in the local optimum. In this case, a huge amount of sample will be drawn to find enough accepted particles, and the new generation will be able to jump out of the local optimum and have different concentrations.

@@

To illustrate this, Two independent run, namely Run A and Run B, were chosen from the experiment of $p=24$ and visualised in Figure \ref{fig:local_modes} and Figure \ref{fig:local_para}. These two runs used identical options for the algorithm and implementation. Figure \ref{fig:local_modes} (a) shows the comparison of the required number of samples, where Run A converged to a local mode. In population 11 Run A took much more sample tires to move away from the local optimum. Figure \ref{fig:local_para} compares the posterior density distribution before (t=9, cyan) and after (t=13, red) the concentration jumping: significant movement of concentrations in posterior density distributions are observed, especially in parameters from ODE 1, 2 and 4. An interning movement is that there are two concentrations in $\kappa_{\Phi\beta}$ and $\mu_\Phi$ when t=9, and the majority of particles are concentrated in the second concentration with higher values; when t=13, the second concentration distribution disappears and all particles are concentrated near zero.

@@

By looking into the code, some reasons can be found. There exist some `cheap' particles. If at the first time points, a variable reached unacceptably high value (regarded as \verb|Inf| in \verb|Python|), the ODE solver cannot perform further integration but result in the simulated data full of \verb|NaN|s. Such particles cost less time than a usual particle (where values at 120 data points are explicitly calculated and then compared with observed data). Local mode, e.g. generation 11, might take a large amount of these `cheap' particles, thus it is `efficient' regarding the per-second performance. Also, generation 1 draws samples directly from prior without fitting and using kernels, so it is much faster. The huge amount of time observed in generation 2 is believed to be a result of some initialisation steps in the ABC SMC algorithm.

To conclude, the performance experiments revealed reasonable scaling-up behaviour when the \verb|Python| processes did not exceed the number of physical cores. Local modes lead to high variance in total required number of samples, and the relationship between required number of sample numbers and required execution time are not ideally linear; they all made further detailed analysis difficult.

Additionally, by observing CPU frequency and load curve in-time while running the program, we found that the parallelised part is limited to the sampling, calculation and criterion test process. Some reduction calculations and preparation work for the next generation (e.g. calculation of the new epsilon values, normalisation of data, fit of multivariate kernel) are serial. The database I/O is also serial.  It is noticed that the program takes no advantage of graphics card, as \verb|multiprocessing| along can only exploit the resources inside CPU.
